Demes: a standard format for demographic models

### 2022-08-03 response-to-reviewers.tex

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication, for your
positive comments and in particular for the helpful and insightful reviews.

The changes that we have made mostly reflect an attempt to better
explain the purpose of Demes, its scope, and the rationale for
key design decisions. While the changes to the text
look fairly extensive (see the attached latex-diff), they are mostly
a function of moving text from the appendix to the main text, and
removing redundancy.

Specifically:

\begin{itemize}
\item We compressed the text and improved the presentation by removing
the Parsers appendix (there was little extra information that was not
already in the existing main text section).
\item Laurent made some good and fair points, but which were mainly
(we believe) reflecting our failure at communicating
the underlying rationale for design decisions effectively.
We have rearranged the text, bringing some content that was in the
appendix into the main text to try and address this up front
and avoid similar misunderstandings.
\end{itemize}

We have responded to all comments point-by-point, below.

@@

\begin{point}
I don't have much to add to this manuscript other than a few comments that I
hope will help answer a few questions that other readers could have. My main
comment below is that an example of the YAML format should be included in the
main text.
\end{point}
\begin{reply}
We have addressed each of the comments in turn below.
\end{reply}

\begin{point}
I usually like when there is a link to the software directly in the abstract
 in those kind of papers, it avoids scrolling through it as links are
sometimes in different places in those type of manuscripts.
\end{point}
\begin{reply}
We agree that this is a useful practice.
In our case there are several software packages that are directly
and indirectly related to the Demes format as summarized in Table 1,
so we have instead added a link to the specification documentation in the abstract.
This documentation includes a tutorial introduction for users to learn
how to write a Demes YAML file, and a detailed specification that developers
may consult when implementing Demes support in their own sofware.
\end{reply}

\begin{point}
I suggest including the example of the YAML format in the main text.
This is too sad that the center piece of this paper has been relegated to the appendix :(
Show us how easy it is to understand the model from the YAML format!
Maybe FigA2 is a bit too long, but FigA1 A and B could work well.
\end{point}
\begin{reply}
Thank you for this excellent suggestion.
We have moved the example model in Figure A1.A (the HDM YAML) and
Figure A1.B (the model diagram) into the main text as Figures 1A and B,
while leaving the MDM form of the same model in the Appendix.
We have also added a section to the text that decribes the main features
of the model in relation to the YAML document.
\end{reply}

\begin{point}
A fair question that could be asked by any potential user could be: Are there
  versions to this data format? There is no version argument in the YAML file,
what if the format changes in the future? How would my model format would be
recognized by the parser? I get that it is designed with stability in mind and
it is done well, but we've seen formats change for other things through time
(e.g. in bioinformatics) to adapt to new usage.
\end{point}
\begin{reply}
We have had an issue open for this for some time (\url{https://github.com/popsim-consortium/demes-spec/issues/20}).
The current plan is to add a variable for this when the spec hits version 1.0, which will coincide
with this publication.
Files omitting a spec version will assume a default value of 1.0.
\end{reply}

\begin{point}
There might be the need to specify how the initialization time of the whole
  model is dealt with. This is kind of straightforward when thinking about
coalescent models for people used to it. But not for forward-time simulations.
Especially that in your examples there are no start\_time for the first
population. This might be hard for some users to wrap their head around.
\end{point}
\begin{reply}
Thank you for bringing this omission to our attention.
It's important that different forward simulators use the same strategy
for converting from times in the past (as used by Demes) into their
corresponding values when using a forwards-time convention.
The technique is to find the oldest non-infinite time in the model
and use this as the forwards-time reference point, suitably adjusted
for a burn-in. This same approach is used by DemesDraw for the model diagrams.
We have added a section to the specification detailing how this
conversion should be done
(\url{https://github.com/popsim-consortium/demes-spec/pull/159}),
and added a brief description to the ``Time units'' section in the manuscript's
appendix.
\end{reply}

\begin{point}
p2 you discuss the defaults of CLI and input parameter formats but not of
  APIs. It could be said that APIs require some learning (of both the interface
and the given programming language for users not used to it). Learning a new
API each time you change simulator is also time consuming.
\end{point}
\begin{reply}
We have updated the text to include this point.
\end{reply}

\begin{point}
p2 ``YAML format" you could specify that it is a data serialization language
\end{point}
\begin{reply}
We have updated the text to include this point.
\end{reply}

\begin{point}
p12 ``time units is in generations" you could precise it is the default value here, if it is?
\end{point}
\begin{reply}
There is no default value for the time units. This must be specified explicitly.
We mention that the time units field is mandatory in the new Example section
that describes the new Figure 1.
\end{reply}

\begin{point}
p12 ``such as years, accompanied by the generation time". Might be worth
  detailing that it is needed as simulators work with generations and not
years, so there needs to be some kind of conversion.
\end{point}
\begin{reply}
We have updated the text to include this point.
\end{reply}

\begin{point}
p12 ``function defined by start and end sizes" You should detail here what are
  the models of population expansion available and what is the default. Linear,
exponential, etc.?
\end{point}
\begin{reply}
We have updated the text to clarify that only the exponential population
size function is currently supported, but the specification may be updated
to include more.
\end{reply}

\begin{point}
p12 ``S are born via [...]" missing that S is a proportion
\end{point}
\begin{reply}
Fixed.
\end{reply}

\begin{point}
p13-14 ``Other implicit values [...]" until end of paragraph, this part with
  the example could be worth integrating into the main text as this is a major
thing to understand about the data model in my opinion.
\end{point}
\begin{reply}
We have rephrased this, and illustrated in the new ``Example'' section.
\end{reply}

@@

\begin{point}
Section A3: you could also detail that having common parsers allows for
  consistent and informative error messages when it comes to missing values or
issues in formatting.
\end{point}
\begin{reply}
We have updated the Parsers main text section to make this point.
\end{reply}

\begin{point}
p15 what's the difference between ``genome annotations" and ``sequence annotations"?
\end{point}
\begin{reply}
Good point; we have deleted the ``sequence annotations''.
\end{reply}

@@

\subsection*{Random thoughts I had while reading the paper, to discuss (or
not)}

\begin{point}
I wondered if such format couldn't be used as a way to increase (spoken)
  language inclusivity. I guess it would be fairly easy to have dictionaries
including multiple translations for each keyword of the format in the parser.
That way you could write your model in Japanese or Spanish and it could still
be read without issue.
\end{point}
\begin{reply}
The field names in a YAML file are variables, not words \textit{per se}.
Thus, allowing different inputs for the field names is like having code trying
to understand different names for the same piece of data.
While it would be useful to allow users to generate files with variable names
that are not English words, and in character sets other than Latin, those files would
require additional conversion into the variable/value format defined by \textit{demes}.
We feel that such conversion is beyond the scope of this project.
\end{reply}

\begin{point}
Do you think such plain text format also provide a better substrate for
versionning demographic models? Could be mentioned.
\end{point}
\begin{reply}
While plain-text formats are ideally suited for use with version control
systems such as git, almost all competing/custom formats are also
plain text and thus share this property. However, Demes models are more
clearly separated from code and CLI directives, which may make model
changes more obvious when reviewing change requests such as git pull requests.
\end{reply}

@@

\reviewersection

\begin{point}
In their paper entitled ``Demes: a standard format for demographic models",
Gower et al. propose a new way to describe (historical) demographic models of
populations.

### 2022-08-03 response-to-reviewers.tex

Please see the specification
(\url{https://popsim-consortium.github.io/demes-spec-docs/main/specification.html#defaults})
and tutorial
(\url{https://popsim-consortium.github.io/demes-spec-docs/main/tutorial.html#setting-defaults})
for the detailed documentation on this.
\end{reply}

\begin{point}
p. 14. Parsers. It is a good idea to provide HDM to MDM parsers, but I note
  that each program reading an MDM will also implement its own parser to read
it and translate it to program specific parameters. Again, it would make more
sense to me to have a single format and no parser. Rather some parser making
sure that the demographic model makes sense could be useful: i.e. checking that
population ancestors have all been defined, or that end\_times are compatible
with start\_time.
\end{point}
\begin{reply}
It's not necessarily true that a program that accepts the MDM will need to
write its own Demes (MDM) parser. For example, msprime uses the \texttt{demes}
Python module to accept demographic models in Demes format. In effect,
what msprime does is take the object model provided by the parser
which is identical to an MDM document. Msprime then processes this document,
converting it into its own internal Demography format. Once the conversion
into msprime's internal format is complete, \emph{this} object
model is validated using precisely the same code as if the model
had been produced by msprime's own demography APIs.
The only validation that msprime does on the Demes MDM document
is to check for parameters that it doesn't support (like selfing and cloning
rates).

Thus, from msprime's perpective, there is no difference between what we
have currently (the HDM converted to MDM via the parser) and what
you propose (the MDM validated by the parser).

If the HDM did not exist, we would still require parsers because validating
the full MDM is a lengthy and tedious task.
A key goal that we set was for the ecosystem to be ``batteries-included''
from the outset, which is why we have implemented several high-quality
parsers (Python, C, Rust, and Julia) which should
cover the vast majority of use-cases.
\end{reply}

### 2022-08-03 paper.tex
fix typo
very minor changes

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being proposed and tested.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labor intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in several simulators and inference
methods.
An introduction to the file format and a detailed specification are available at:\\
\url{https://popsim-consortium.github.io/demes-spec-docs/}.
\end{abstract}

@@

\subsection*{Population genetics model}
For inference and simulation software to meaningfully interoperate there
must be a shared understanding of what a demographic model \emph{is}.
Population genetics is a large field, and rather than attempting to
capture all possible within- and between- population processes, we have
instead adopted a pragmatic approach of identifying a common set
of assumptions shared by many methods. We outline the processes
and assumptions briefly here and in
Appendix~\ref{sec:appendix-pop-gen-model}.

@@

Structurally, the HDM encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
For values that repeat
across fields, the ``defaults'' mechanism may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Size values are inherited naturally following the
the document. Size values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided (either directly, or via a defaults section),
it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size (or it must be provided in a defaults section).

@@

While the HDM is designed for human readability and
conciseness, the underlying data model suitable for software implementation
(the Machine Data Model, or MDM)
is redundant and exhaustive.
Translation from the HDM to the MDM
requires resolving hierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as input.
The Demes specification precisely defines the required behavior of parsers,
and we provide a reference implementation written in Python to resolve any
potential ambiguities, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (Table~\ref{tab:software})
providing a solid foundation for the software ecosystem.
By maintaining high-quality Demes parsers available as libraries,
we ensure consistency across simulation and inference
software.
Having common parsers also benefits users by providing
consistent and informative error messages for missing values or
issues in formatting.

@@

The model is therefore limited to features that we can expect many different
demographic inference and simulation methods to share. The specification only
describes demographic features at the population level. Features of genome
biology are out of scope, including mutation and recombination rates,
genome annotations, ploidy, and so on. Selection and dominance models are
absent, as discussed in Appendix~\ref{sec:appendix-pop-gen-model}.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modelled in the specification, such as interpreting population sizes as
carrying capacities, implementations of hard selection, or layering more
complicated mating or spatial structure. The Demes specification
is intended to provide a basic model that can be elaborated on where necessary.

Demes is not a standard population genetic simulation specification,
although it could be \emph{part} of one.
Since the standard is based on JSON, and JSON documents can be
arbitrarily nested, it is possible to imagine a simple specification
of genome features such as mutation and recombination rates
in which the demography is defined by an embedded Demes specification.
Features of the simulation specification (such as defining
the time and location of samples) can then \emph{refer to} the Demes
model. This design, in which we embed the demographic model
\emph{within} a larger specification rather than adding arbitrary
and unrelated complexities \emph{to} the demography
is an essential simplification and separation of duties.

>>>>

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being proposed and tested.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labor intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers
in multiple languages including Python and C,
and showcase initial support in several simulators and inference
methods.
An introduction to the file format and a detailed specification are available at:\\
\url{https://popsim-consortium.github.io/demes-spec-docs/}.
\end{abstract}

@@

\subsection*{Population genetics model}
For inference and simulation software to meaningfully interoperate there
must be a shared understanding of what a demographic model \emph{is}.
Population genetics is a large field, and rather than attempting to
capture all possible within- and between-population processes, we have
instead adopted a pragmatic approach of identifying a common set
of assumptions shared by many methods. We outline the processes
and assumptions briefly here and in
Appendix~\ref{sec:appendix-pop-gen-model}.

@@

Structurally, the HDM encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
For values that repeat
across fields, the ``defaults'' mechanism may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Size values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided (either directly, or via a defaults section),
it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size (or it must be provided in a defaults section).

@@

While the HDM is designed for human readability and
conciseness, the underlying data model suitable for software implementation
(the Machine Data Model, or MDM)
is redundant and exhaustive.
Translation from the HDM to the MDM
requires resolving hierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as input.
The Demes specification precisely defines the required behavior of parsers,
and we provide a reference implementation written in Python to resolve any
potential ambiguities, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python, C, Rust, and Julia languages (Table~\ref{tab:software})
providing a solid foundation for the software ecosystem.
By maintaining high-quality Demes parsers available as libraries,
we ensure consistency across simulation and inference
software.
Having common parsers also benefits users by providing
consistent and informative error messages for missing values or
issues in formatting.

@@

The model is therefore limited to features that we can expect many different
demographic inference and simulation methods to share. The specification only
describes demographic features at the population level. Features of genome
biology are out of scope, including mutation and recombination rates,
genome annotations, ploidy, and so on. Selection and dominance models are
absent, as discussed in Appendix~\ref{sec:appendix-pop-gen-model}.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modeled in the specification, such as interpreting population sizes as
carrying capacities, implementations of hard selection, or layering more
complicated mating or spatial structure. The Demes specification
is intended to provide a basic model that can be elaborated on where necessary.

Demes is not a standard population genetic simulation specification,
although it could be \emph{part} of one.
Since the standard is based on JSON, and JSON documents can be
arbitrarily nested, we can imagine a simple specification
of genome features such as mutation and recombination rates
in which the demography is defined by an embedded Demes specification.
Features of the simulation specification (such as defining
the time and location of samples) can then \emph{refer to} the Demes
model. This design, in which we embed the demographic model
\emph{within} a larger specification rather than adding arbitrary
and unrelated complexities \emph{to} the demography
is an essential simplification and separation of duties.

### 2022-08-03 software-table.tex
Linkify urls.
Also Americanize software-table.tex.
very minor changes

\texttt{demes-c} &
    A C library for parsing Demes YAML descriptions.
    (\url{https://github.com/grahamgower/demes-c}). \\

\texttt{demesdraw} &
    A Python library for visualising Demes models
    (\url{https://github.com/grahamgower/demesdraw}). \\

>>>>

\texttt{demes-c} &
    A C library for parsing Demes YAML descriptions
    (\url{https://github.com/grahamgower/demes-c}). \\

\texttt{demes-rust} &
    A Demes parser in Rust
    (\url{https://github.com/molpopgen/demes-rs}). \\

\texttt{demes-julia} &
    A parser in Julia
    (\url{https://github.com/apragsdale/Demes.jl}). \\

\texttt{demesdraw} &
    A Python library for visualizing Demes models
    (\url{https://github.com/grahamgower/demesdraw}). \\
